Well-quasi-orderings on word languages

### 2024-08-07 src/infixes.tex
writing proofs

% lang: en-US

\section{Infixes}
\label{infixes:sec}

@@

\begin{theorem}
    \label{thm:infix-embedding}
    Let $(X, \preceq)$ be a quasi-ordered set.
    Then the following are equivalent
    \begin{itemize}
        \item $X$ embeds into $(\Sigma^*, \preceq_{\text{infix}})$
        \item $X$ is countable, and for every $x \in X$,
            $\{y \in X \mid y \preceq x\}$ is finite.
    \end{itemize}

    Furthermore, if $X$ is recursively enumerable and $\preceq$ decidable,
    then
    the embedding is effective.
\end{theorem}

As a consequence of \cref{thm:infix-embedding}, we cannot replay proofs of
\cref{prefixes-suffixes:sec}, and will actually need to leverage some
regularity of the languages to obtain a characterization of well-quasi-ordered
languages under the infix relation. Let us first play this game for languages
that are recognized by finite automata.

\begin{theorem}
    \label{thm:infix-finite-automata}
    Let $L \subseteq \Sigma^*$ be a language recognized by a finite automaton.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

@@

\subsection{Amalgamation systems}

@@

@@

\subsection{Monoid equations?}

\begin{itemize}
    \item What are the finite monoids $(M, \cdot)$ such that
    every language recognized by $M$ is well-quasi-ordered by the infix relation?
    \item What are the equations on $M$ and $P \subseteq M$ characterizing 
    the well-quasi-ordering of $L(M, P)$?
\end{itemize}

>>>>

% LTeX: language=en-GB 
\section{Infixes and Regular Languages}
\label{infixes-regular:sec}

@@

Let us recall that an \intro{order embedding} from a quasi-ordered set $(X,
\preceq)$ into a quasi-ordered set $(Y, \preceq')$ is a function $f \colon X
\to Y$ such that for all $x, y \in X$, $x \preceq y$ if and only if $f(x)
\preceq' f(y)$. When such an embedding exists, we say that $X$ \reintro{embeds
into} $Y$.

We say that a quasi-ordered set $(X, \preceq)$ is a \intro{partial ordering}
whenever the relation $\preceq$ is antisymmetric, that is $x \preceq y$ and $y
\preceq x$ implies $x = y$.

\begin{theorem}
    \label{infix-embedding:thm}
    Let $(X, \preceq)$ be a \kl{partially ordered} set,
    and $\Sigma \defined \set{a,b,c}$.
    Then the following are equivalent:
    \begin{enumerate}
        \item \label{infix-embedding-embeds:item} 
            $X$ \kl{embeds into} $(\Sigma^*, \infleq)$,
        \item \label{infix-embedding-count:item}
            $X$ is countable, and for every $x \in X$,
        $\setof{y \in X}{y \preceq x}$ is finite.
    \end{enumerate}
    Furthermore, if $X$ is recursively enumerable and $\preceq$ decidable,
    then
    the \kl{embedding} is computable.
\end{theorem}
\begin{proof}
    Let us first prove that \cref{infix-embedding-embeds:item} implies
    \cref{infix-embedding-count:item}. Let $f \colon X \to \Sigma^*$ be an
    \kl{embedding}. Note that $f$ is injective, because $X$ and $\Sigma^*$
    are \kl{partially
    ordered}. In particular, because $\dwset[\infleq]{w}$ is finite
    for every $w \in \Sigma^*$, then so is $\setof{y \in X}{y \preceq x}$
    for every $x \in X$. Similarly, $X$ must be countable
    because $\Sigma^*$ is.

    Conversely, let us prove that \cref{infix-embedding-count:item} implies
    \cref{infix-embedding-embeds:item}. To that end, let us first define
    inductively a sequence $\seqof{X_n}$ of subsets of $X$ as follows: $X_0$ is
    the set of minimal elements of $X$, and $X_{n+1}$ is the set of minimal
    elements of $X \setminus \bigcup_{i \leq n} X_i$. Because initial segments
    of $X$ are finite, it is clear that $X = \bigcup_{n \in \Nat} X_n$. To
    simplify notations, we will also use $Y_n \defined \bigcup_{i \leq n} X_i$.
    We also let $L_0 \defined a b^* a$, and 
    $L_{n+1} \defined L_n \cup c^{n+1} a^{n+1} b^* a^{n+1} c^{n+1} (L_n c^{n+1})^*$.
    Now, let us inductively define \kl{embeddings} $f_n \colon Y_n \to
    L_n$.

    For the base case, let $\seqof{x_i}[i \in \Nat]$ be an enumeration of
    $X_0$, which is possible because $X$ is countable. We define $f_0 \colon
    X_0 \to \Sigma^*$ as follows: $f_0(x_i) \defined a b^i a$. It is immediate
    that $f_0$ is an \kl{embedding}. For the inductive step, let $\seqof{x_i}[i
    \in \Nat]$ be an enumeration of $X_{n+1}$, and assume that $f_n$ is an
    \kl{embedding} from $Y_n$ to $\Sigma^*$.
    Let us define $D_i \defined \setof{ y \in Y_n }{ y \preceq x_i }$
    for $i \in \Nat$, all of which are finite by assumption.
    We then let for all $i \in \Nat$:
    \begin{equation*}
        f_{n+1}(x_i) \defined 
        \underbrace{c^{n+1} a^{n+1} b^i a^{n+1} c^{n+1} \prod_{y \in D_i} \overbrace{f_n(y)}^{\in L_n} c^{n+1}}_{\in L_{n+1}}
        \quad .
    \end{equation*}
    And let $f_{n+1}(y) \defined f_n(y)$ for all $y \in Y_n$.

    To check that $f_{n+1}$ is an \kl{embedding}, let us first notice that
    words $f_{n+1}(x_i)$ and $f_{n+1}(x_j)$ are incomparable for the \kl{infix
    relation} when $i \neq j$, because the only factor of the form $c^{n+1}
    a^{n+1} b^t a^{n+1} c^{n+1}$ appears at the beginning of these words, and
    precisely encode $i$ (resp. $j$) in the value of $t$. Now, let us consider
    $y \in Y_n$ and $x_i \in X_{n+1}$ such that $y \preceq x_i$. Because $y \in
    D_i$, we have $f_n(y) \infleq f_{n+1}(x_i)$ by construction, and since
    $f_n(y) = f_{n+1}(y)$, we conclude that $f_{n+1}(y) \infleq f_{n+1}(x_i)$. 
    Conversely, assume that
    $f_{n+1}(y) \infleq f_{n+1}(x_i)$. Again, this means
    that $f_n(y) \infleq f_{n+1}(x_i)$. Now, the encoding
    is robust enough so that we can conclude
    $f_{n}(y) \infleq f_{n}(z)$ for some $z \in D_i$.
    By induction hypothesis, $y \preceq z$, hence, $y \preceq x_i$.

    We conclude by noticing that $\bigcup_{n \in \Nat} f_n$ is an \kl{embedding}
    of $X$ into $\Sigma^*$.
\end{proof}

As a consequence of \cref{infix-embedding:thm}, we cannot replay proofs of
\cref{prefixes:sec}, and will actually need to leverage some
regularity of the languages to obtain a characterization of well-quasi-ordered
languages under the infix relation. Let us first play this game for languages
that are recognized by finite automata.

\textbf{TODO: introduce finite automata and refer to the book.}.

\begin{theorem}
    \label{infix-finite-automata:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by a finite automaton.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

In order to prove \cref{infix-finite-automata:thm}, we will perform
some preliminary analysis on the structure of an automaton recognizing a
well-quasi-ordered language under the infix relation.

\begin{lemma}
    Let $A = (Q, \Sigma, \delta, q_0, F)$ be a finite automaton recognizing a
    language $L \subseteq \Sigma^*$. Assume furthermore that
    $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, for all words $u,w,v \in \Sigma^*$, and states
    $q, q_f \in Q \times F$ such that
    $q_0 \xrightarrow{u} q \xrightarrow{w} q \xrightarrow{v} q_f$,
    one of the following holds:
    \begin{itemize}
        \item There exists $k \in \Nat$
            such that $u$ is a \kl{suffix} of $w^k$,
        \item There exists $k \in \Nat$
            such that $v$ is a \kl{prefix} of $w^k$.
    \end{itemize}
\end{lemma}

\begin{lemma}
    Let $A = (Q, \Sigma, \delta, q_0, F)$ be a finite automaton recognizing a
    language $L \subseteq \Sigma^*$. Assume furthermore that
    $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, for all words $w_1, w_2 \in \Sigma^*$, and states
    $q, q_f \in Q \times F$ such that
    $q_0 \rightarrow^* q \xrightarrow{w} q \rightarrow^* q_f$
    and 
    $q_0 \rightarrow^* q \xrightarrow{w_2} q \rightarrow^* q_f$,
    there exists a word $w \in \Sigma^*$
    such that $w_1$ and $w_2$ are powers of $w$.
\end{lemma}

\begin{lemma}
    Let $A = (Q, \Sigma, \delta, q_0, F)$ be a finite automaton recognizing a
    language $L \subseteq \Sigma^*$. Assume furthermore that
    $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, for all words $u, w_1, w_2 \in \Sigma^*$, and states
    $q_1, q_2, q_f \in Q^2 \times F$,
    such that
    $q_0 \rightarrow^* q_1 \xrightarrow{w_1} q_1 \xrightarrow{u} q_2 \xrightarrow{w_2} q_2
    \rightarrow^* q_f$,
    there exists a $w \in \Sigma^*$ such that $w_1$, $w_2$, and $u$ are powers of $w$.
\end{lemma}

We are now ready to prove the main result.

\begin{proofof}{infix-finite-automata:thm}
    Let $w \in L$. Consider a run of the automaton on $w$. If $w$ is too long,
    then there exists a loop, and we can prove that $w = u x^p v$ for some
    short period $x$. By repeating the process for $u$ and $v$, we conclude
    that $w$ has the form $w = u_0 x_0^{p_0} u_1 x_1^{p_1} \cdots u_k x_k^{p_k}
    v_k$. Now, applying the loop merging lemma, we conclude that $x_0 = x_1 =
    \cdots = x_k$. Hence, $w = u_0 x^{p} v_k$. Finally, we know that either
    $u_0$ is a prefix of $x^{p}$ or that $v_k$ is a suffix of $x^{p}$. We
    conclude in particular that $\dwset[\infleq]{ \setof{ u x^p v }{ p \in \Nat
    }}$ is a finite union of \kl{chains}. Because $u,v,x$ are of bounded
    length, we conclude that $L$ itself included in a finite union of
    \kl{chains}, hence is a finite union of \kl{chains}.
\end{proofof}

@@

\section{Infixes and Amalgamation systems}
\label{infixes-amalgamation:sec}

@@

\begin{theorem}
    \label{infix-amalgamation:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by an 
    \kl{amalgamation system}.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

@@

### 2024-08-07 src/introduction.tex
writing proofs

\subparagraph{Outline} In \cref{prefixes:sec}, which is relatively
self-contained, we study the \kl{prefix relation} prove in \cref{prefixes:thm}
the characterization of \kl{well-quasi-ordered} languages by the \kl{prefix
relation}. In \cref{infixes:sec}, we prove that any countable quasi-ordering
with finite initial segments can be embedded into the \kl{infix relation} of a
language (\cref{count-embed-infix:thm}), we obtain the analogue of
\cref{prefixes:thm} for languages recognized by finite automata
(\cref{infix-finite-automata:thm}). Finally, in \cref{amalgamation:sec}, we
introduce \kl{amalgamation systems} and generalize
\cref{infix-finite-automata:thm} to this setting
(\cref{infix-amalgamation:thm}).

>>>>

\subparagraph{Outline} In \cref{prefixes:sec}, which is relatively
self-contained, we study the \kl{prefix relation} prove in \cref{prefixes:thm}
the characterization of \kl{well-quasi-ordered} languages by the \kl{prefix
relation}. In \cref{infixes-regular:sec}, we prove that any countable quasi-ordering
with finite initial segments can be embedded into the \kl{infix relation} of a
language (\cref{infix-embedding:thm}), we obtain the analogue of
\cref{prefixes:thm} for languages recognized by finite automata
(\cref{infix-finite-automata:thm}). Finally, in \cref{infixes-amalgamation:sec}, we
introduce \kl{amalgamation systems} and generalize
\cref{infix-finite-automata:thm} to this setting
(\cref{infix-amalgamation:thm}).

### 2024-08-07 src/metadata.tex
writing proofs

>>>>

\newcommand{\makeabstract}{
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself
    well-quasi-ordered. This seminal result by Higman is a cornerstone
    of the theory of well-quasi-orderings and has found numerous
    applications in computer science. However, this result is based on a
    specific choice of ordering on words, the (scattered) subword
    ordering. In this paper, to what extent other natural orderings
    (prefix, suffix, infix) on words can be used to derive Higman-like
    theorems. More specifically, we are interested in characterizing
    \emph{languages} of words that are well-quasi-ordered under these
    orderings. We show that a simple characterization is possible for
    the prefix and suffix orderings, and that under extra regularity
    assumptions, this also lifts to the infix ordering.
\end{abstract}
}

### 2024-08-07 src/prefixes.tex
writing proofs

% LTeX: language=en-GB 
\section{Prefixes and Suffixes}
\label{prefixes-suffixes:sec}

@@

Notice that the tree of prefixes is finitely branching. Let us now
observe how antichains in the prefix relation can be witnessed
by infinite branches in the tree of prefixes.

@@

\begin{theorem}
    A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the
    \kl{prefix relation} if and only if $L$ is a union of \kl{chains}.
\end{theorem}
\begin{proof}
    Assume that $L$ is a finite union of \kl{chains}.
    Because the \kl{prefix relation} is \kl{well-founded},
    and that finite unions of \kl{chains} have finite \kl{antichains},
    we conclude that $L$ is \kl{well-quasi-ordered}.

    Conversely, assume that $L$ is not a finite union of \kl{chains}. Then
    there exists an infinite \kl{antichain branch} $B$ for $L$. Indeed
    \textbf{todo}. In particular, $L$ is not \kl{well-quasi-ordered}.
\end{proof}

@@

\begin{corollary}
    \textbf{false: some chains can be finite}
    Let $L \subseteq \Sigma^*$ be a language that is \kl{well-quasi-ordered} by
    the \kl{prefix relation}. Then, there exists a finite $k \in \Nat$ such that
    the
    \kl{maximal order type} of $L$ is $k \cdot \omega$,
    the \kl{ordinal height} of $L$ is $\omega$, and its
    \kl{ordinal width} is $k$.
\end{corollary}

>>>>

% LTeX: language=en-GB 
\section{Prefixes and Suffixes}
\label{prefixes:sec}

@@

Because the alphabet $\Sigma$ is assumed to be finite, the \kl{tree of
prefixes} is finitely branching. Let us now observe how \kl{antichains} in the
prefix relation can be witnessed by infinite branches in the \kl{tree of
prefixes}.

@@

\begin{theorem}
    \label{prefixes:thm}
    A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the
    \kl{prefix relation} if and only if $L$ is a union of \kl{chains}.
\end{theorem}
\begin{proof}
    Assume that $L$ is a finite union of \kl{chains}.
    Because the \kl{prefix relation} is \kl{well-founded},
    and that finite unions of \kl{chains} have finite \kl{antichains},
    we conclude that $L$ is \kl{well-quasi-ordered}.

    Conversely, assume that $L$ is \kl{well-quasi-ordered} by the \kl{prefix
    relation}. Let us define $S$ the set of words $w$ such that there exists
    two words $wu$ and $wv$ both in $L$ that are not comparable for the
    \kl{prefix relation}. Assume by contradiction that $S$ is infinite.
    Then, $S$ equipped with the \kl{prefix relation} is an infinite
    tree with finite branching, and therefore contains an infinite
    branch, which is by definition an \kl{antichain branch} for $L$.
    This contradicts the assumption that $L$ is \kl{well-quasi-ordered}.
    Now, consider $w$ be a maximal element for the \kl{prefix ordering}
    in $S$. By definition, all words in $L$ that contain $w$ as a prefix
    must be comparable for the \kl{prefix relation}. Therefore
    $\upset[\prefleq]{w} \cap L$ is a \kl{chain} for the \kl{prefix relation}.
    In particular, letting $S_{\max}$ be the set of all maximal elements
    of $S$,
    we conclude that 
    \begin{equation*}
        L \subseteq S \cup \bigcup_{w \in S_{\max}} \upset[\prefleq]{w} \cap L
        \quad .
    \end{equation*}
    Hence, that $L$ is finite union of \kl{chains}.
\end{proof}

@@

\begin{corollary}
    Let $L \subseteq \Sigma^*$ be an infinite language that is \kl{well-quasi-ordered} by
    the \kl{prefix relation}. Then, there exists finite numbers $k,\ell \in \Nat$ such that
    the
    \kl{maximal order type} of $L$ is $k \cdot \omega$,
    the \kl{ordinal height} of $L$ is $\omega$, and its
    \kl{ordinal width} is $\ell$.
\end{corollary}
\begin{proof}
    \textbf{todo.}
\end{proof}
